What's in a name? Disambiguation strategies for relational descriptions

### 2023-07-12 2023 Disambiguation/Disambiguation.tex

The intended term is Term~\ref{eqn:disambGood} here.
Its meaning is different from that of Term~\ref{eqn:disambBad}, since the term $\epsilon_{t_3};\typecomp{1_{\mathrm{Event}}};\epsilon_{t_3}$ can contain pairs in $\mathrm{Event}$ that $\epsilon_{t_0};\typecomp{1_{\mathrm{Task}}};\epsilon_{t_0}$ does not.
We could argue that the user must specify more type information here, since the meanings for these terms differ.
However, the term can be interpreted as one that does not use sub-typing in any meaningful way.
Since the purpose of a disambiguation system is to reduce verbosity, we should prefer the intended term over Term~\ref{eqn:disambBad}.
A criterium that distinguishes the two terms is that the intended disambiguation uses $\epsilon_{t_0}$ throughout, which will be the preferred term if we prefer $\epsilon_{A*B}$'s that do not employ sub-typing (that is: those where $A=B$).

What if the user writes $\typecomp{\typetyped{\typeident}{\mathrm{Event*Event}}} \typeunion \typecomp{\left(\mathrm{before} \typeinter \typeconv{\mathrm{before}}\right)}$, a term forces the use of sub-typing? Now possible disambiguated terms include:
\begin{align}
\epsilon_{t_3};\overline{\typetyped{1_{\mathrm{Event}}}{t_3}};\epsilon_{t_3} &\cup \epsilon_{t_3};\overline{\left(\epsilon_{t_1};\mathrm{before}_{t_0};\epsilon_{t_2} \cap \epsilon_{t_1};\conv{\mathrm{before}_{t_0}};\epsilon_{t_2} \right)};\epsilon_{t_3}\label{eqn:disambOption}\\
\epsilon_{t_3};\overline{\typetyped{1_{\mathrm{Event}}}{t_3}};\epsilon_{t_3} &\cup \epsilon_{t_1};\overline{\left(\epsilon_{t_0};\mathrm{before}_{t_0};\epsilon_{t_0} \cap \epsilon_{t_0};\conv{\mathrm{before}_{t_0}};\epsilon_{t_0} \right)};\epsilon_{t_2}\label{eqn:disambOld}\\
\shortintertext{\flushright{ with ${t_0} = \mathrm{Task*Task}$, ${t_1} = \mathrm{Event*Task}$, ${t_2} = \mathrm{Task*Event}$, and ${t_3} = \mathrm{Event*Event}$}}\nonumber{}
\end{align}

Observe that again, the two terms have different semantics, due to the complement being computed over a different type.
In previous publications on the Ampersand type system, where sub-typing did not play a role, Term~\ref{eqn:disambOld} would be chosen over Term~\ref{eqn:disambOption}.
The rational here is that types are computed bottom up.
However, we have observed that sub-typing around complements can be error-prone, and the Term~\ref{eqn:disambOption} might well have been intended.
We therefore argue that an error should be raised here.
The algorithm presented can be configured either way: accepting the term as Term~\ref{eqn:disambOld}, or throwing an error.
%As a consequence, we argue that it is desirable to require more information from the user here.
%terms like $\typecomp{\typetyped{\typeident}{\mathrm{Event*Event}}} \typeunion \typecomp{\left(\mathrm{before} \typeinter \typeconv{\mathrm{before}}\right)}$ are

@@

Aside from type errors, we can use $p$ to find ambiguous terms:
If there is a sub-term $\Xi$ of $e$ that can be replaced by both $x\in \Xi$ and $y\in Xi$ for some $x \neq y$, then $p(e')$ and $p(e'')$ are nonempty for $e'$ and $e''$ equal to $e$ under the substitution of $\Xi$ for $\{x\}$ and $\{y\}$ respectively.
Furthermore, we can use $p$ to find out whether a negation can have multiple types.

A second operation at our disposal is one that selects preferred $\mathcal{R}$-terms for each $\Xi$ as a way to avoid ambiguous terms.
This operation fulfils our requirement that we should prefer some terms over others.
Let $f : \powerset{\mathcal{R}}\to\powerset{\mathcal{R}}$ be an idempotent function (a function such that $f(f(x))=f(x)$ for all $x$) that returns a subset of its argument, indicating which elements would be preferred.
In particular, choosing $f(\Xi) = \left\{\epsilon_{A*B} \mid \epsilon_{A*B}\in \Xi \wedge A=B\right\}$ would have us pick $\epsilon_{A*B}$ relations with $A=B$.
Given such a function $f$, we say that $e'$ is more preferable to $e$ if $p(e')\neq \emptyset$ and $e'$ can be obtained from $e$ by substituting some $\Xi$ terms for $f(\Xi)$.
We say that $e$ is {\emph maximally preferable} if $p(e)\neq \emptyset$ and for every occurrence of a $\Xi$-term in $e$, substituting it for $f(\Xi)$ would either leave $e$ unchanged (as $\Xi=f(\Xi)$ or result in a term $e'$ such that $p(e')=\emptyset$.
Terms that are more preferable to $e$ form a lattice\footnote{Proof: `more preferable' is the subset relation for the set of $\Xi$-term occurrences that have been changed wrt $e$.}, and we call the least upper bound to all maximally preferable terms the term that is generally preferred to $e$ under $f$.
We believe\footnote{Sketch: Create a maximally preferable term using an eager algorithm, then mark all the $\Xi$-terms that are not in such maximally preferable terms as not in the generally preferred term. Keep going starting from $\Xi$-terms that are known not to be in the generally preferred term. Each iteration a new term that is not in the generally preferred term should be discovered: if not, we are done and the remaining terms are those that can be replaced in the generally preferred term.} we can find the generally preferred term in polynomial time.

>>>>

The intended term is Term~\ref{eqn:disambGood} here.
Its meaning is different from that of Term~\ref{eqn:disambBad}, since the term $\epsilon_{t_3};\overline{1_{\mathrm{Event}}};\epsilon_{t_3}$ can contain pairs in $\mathrm{Event}$ that $\epsilon_{t_0};\overline{1_{\mathrm{Task}}};\epsilon_{t_0}$ does not.
We could argue that the user must specify more type information here, since the meanings for these terms differ.
However, the term can be interpreted as one that does not use sub-typing.
Since the purpose of a disambiguation system is to reduce verbosity, we should prefer the intended term over Term~\ref{eqn:disambBad}.
A criterium that distinguishes the two terms is that the intended disambiguation uses $\epsilon_{t_0}$ throughout, which will be the preferred term if we prefer $\epsilon_{A*B}$'s that do not employ sub-typing (that is: those where $A=B$).

What if the user writes $\typecomp{\typetyped{\typeident}{\mathrm{Event*Event}}} \typeunion \typecomp{\left(\mathrm{before} \typeinter \typeconv{\mathrm{before}}\right)}$, a term that forces the use of sub-typing? Now possible disambiguated terms include:
\begin{align}
\epsilon_{t_3};\overline{\typetyped{1_{\mathrm{Event}}}{t_3}};\epsilon_{t_3} &\cup \epsilon_{t_3};\overline{\left(\epsilon_{t_1};\mathrm{before}_{t_0};\epsilon_{t_2} \cap \epsilon_{t_1};\conv{\mathrm{before}_{t_0}};\epsilon_{t_2} \right)};\epsilon_{t_3}\label{eqn:disambOption}\\
\epsilon_{t_3};\overline{\typetyped{1_{\mathrm{Event}}}{t_3}};\epsilon_{t_3} &\cup \epsilon_{t_1};\overline{\left(\epsilon_{t_0};\mathrm{before}_{t_0};\epsilon_{t_0} \cap \epsilon_{t_0};\conv{\mathrm{before}_{t_0}};\epsilon_{t_0} \right)};\epsilon_{t_2}\label{eqn:disambOld}\\
\shortintertext{\flushright{ with ${t_0} = \mathrm{Task*Task}$, ${t_1} = \mathrm{Event*Task}$, ${t_2} = \mathrm{Task*Event}$, and ${t_3} = \mathrm{Event*Event}$}}\nonumber{}
\end{align}

Observe that again, the two terms have different semantics, due to the complement being computed over a different type.
In previous publications on the Ampersand type system, where sub-typing did not play a role, Term~\ref{eqn:disambOld} would be chosen over Term~\ref{eqn:disambOption}.
The rationale here is that types are computed bottom up.
However, we have observed that sub-typing around complements can be error-prone, and the Term~\ref{eqn:disambOption} might well have been intended.
We therefore argue that an error should be raised here.
The algorithm presented can be configured either way: accepting the term as Term~\ref{eqn:disambOld}, or throwing an error.
%As a consequence, we argue that it is desirable to require more information from the user here.
%terms like $\typecomp{\typetyped{\typeident}{\mathrm{Event*Event}}} \typeunion \typecomp{\left(\mathrm{before} \typeinter \typeconv{\mathrm{before}}\right)}$ are

@@

Aside from type errors, we can use $p$ to find ambiguous terms:
If there is a sub-term $\Xi$ of $e$ that can be replaced by both $x\in \Xi$ and $y\in \Xi$ for some $x \neq y$, then $p(e')$ and $p(e'')$ are nonempty for $e'$ and $e''$ equal to $e$ under the substitution of $\Xi$ for $\{x\}$ and $\{y\}$ respectively.
Furthermore, we can use $p$ to find out whether a negation can have multiple types.

A second operation at our disposal is one that selects preferred $\mathcal{R}$-terms for each $\Xi$ as a way to avoid ambiguous terms.
This operation fulfils our requirement that we should prefer some terms over others.
Let $f : \powerset{(\mathcal{R})}\to\powerset{(\mathcal{R})}$ be an idempotent function (a function such that $f(f(x))=f(x)$ for all $x$) that returns a subset of its argument, indicating which elements would be preferred.
In particular, choosing $f(\Xi) = \left\{\epsilon_{A*B} \mid \epsilon_{A*B}\in \Xi \wedge A=B\right\}$ would have us prefer $\epsilon_{A*B}$ relations with $A=B$.
Given such a function $f$, we say that $e'$ is more preferable to $e$ if $p(e')\neq \emptyset$ and $e'$ can be obtained from $e$ by substituting some $\Xi$ terms for $f(\Xi)$.
We say that $e$ is {\emph maximally preferable} if $p(e)\neq \emptyset$ and for every occurrence of a $\Xi$-term in $e$, substituting it for $f(\Xi)$ would either leave $e$ unchanged (as $\Xi=f(\Xi)$) or result in a term $e'$ such that $p(e')=\emptyset$.
Terms that are more preferable to $e$ form a lattice\footnote{Proof: `more preferable' is the subset relation for the set of $\Xi$-term occurrences that have been changed wrt $e$.}, and we call the least upper bound to all maximally preferable terms the term that is generally preferred to $e$ under $f$.
We believe\footnote{Sketch: Create a maximally preferable term using an eager algorithm, then mark all the $\Xi$-terms that are not in such maximally preferable terms as not in the generally preferred term. Keep going starting from $\Xi$-terms that are known not to be in the generally preferred term. Each iteration a new term that is not in the generally preferred term should be discovered: if not, we are done and the remaining terms are those that can be replaced in the generally preferred term.} we can find the generally preferred term in polynomial time.
